MassWateR: Improving Quality Control, Analysis, and Sharing of Water Quality Data

### 2023-08-20 manu-draft.qmd
more formatting, figures from code examples now include captions and citations

execute: 
  echo: false
  warning: false
  message: false
---

```{r}
#| include: false

library(dplyr)
library(here)

@@

@@

The long-term protection and restoration of aquatic resources depends on robust monitoring data; data that require systematic quality control and analysis tools. The *MassWateR* package facilitates quality control, analysis, and data sharing for discrete surface water quality data collected by monitoring programs of various size and technical capacity. The tools were developed to address regional needs for programs in Massachusetts, USA, but the principles and outputs can be applicable to monitoring data collected anywhere.  Users can create quality control reports, perform outlier analyses, and assess trends by season, date, and site for more than 40 parameters. Users can also prepare data for submission to the USEPA Water Quality Exchange, thus sharing data to the largest water quality database in the United States. The automated and reproducible workflow offered by *MassWateR* is expected to increase the quantity and quality of publicly available data to support the management of aquatic resources.

@@

![Workflow demonstrating how a user could engage with the *MassWateR* package. WQX: Water Quality Exchange; QC: Quality Control.](figs/workflow.png){#fig-workflow width=100% fig-alt="Workflow diagram showing four ways to engage with *MassWateR*"}

@@

![Pseudocode demonstrating the iterative process of importing a required data file for *MassWateR*.  All read functions import an Excel file from the user-specified path (e.g., `respth`) and the imported file is then passed to a check function.  The function exits if an error is encountered, allowing the user to manually fix the identified error and then import again.  After all checks are passed, a formatting function is applied to correct minor issues (e.g., standardize date format as YYYY-MM-DD) and the final data object is returned (e.g., `resdat`).](figs/importflow.png){#fig-importflow fig-alt="Diagram showing pseudocode for importing and fixing a data file with *MassWateR*" width=100%}

@@

The `anlzMWRoutlier()` function uses the results file and data quality objectives for accuracy to plot potential outliers for a parameter by month.  The accuracy file is used to automatically identify the y-axis scaling (as arithmetic or log) and to replace concentrations with appropriate values for those labelled as beyond detection. Outliers are defined using the standard definition of 1.5 times the interquartile range (the 25th to 75th percentile) of a parameter and can be visually identified as points above or below the whiskers in the boxplots.  The station name for a point is also shown for identification.  The `param` argument specifies the water quality parameter to evaluate and the `group` argument specifies how the boxplots are grouped (by year, month, week, or station).  In the following example, three stations are identified as potential outliers for the respective month.  Note that the outliers may differ slightly by the grouping choice as they are specific to each value on the x-axis.

```{r}
#| fig-height: 4
#| fig-width: 8
#| echo: true
#| out-width: '95%'
#| fig-align: 'center'
anlzMWRoutlier(fset = fsetls, param = "DO", group = "month")
```

@@

![](figs/qcreview1.png)

![](figs/qcreview2.png)

@@

Seasonal trends can be evaluated using the `anlzMWRseason()` function that summarizes results for a single parameter using boxplots or barplots with seasonal groups assigned to months or weeks of the year. Boxplots or barplots can also include jittered points of the observations on top or only the jittered points can be shown.  Below demonstrates a jittered boxplot of dissolved oxygen observations by month.

```{r}
#| fig-height: 4
#| fig-width: 8
#| echo: true
#| out-width: '95%'
#| fig-align: 'center'
anlzMWRseason(fset = fsetls, param = "DO", thresh = "fresh", 
  group = "month", type = "jitterbox")
```

@@

The `anlzMWRdate()` function plots results as time series on the date the samples or measurements were collected.  This information can be used to evaluate how samples have changed at individual sites over time or as aggregate samples across sites.  In the former case, individual points and lines are used for each site with appropriate labels. For the latter case, sites are aggregated and summary statistics are shown as the mean with optional 95% confidence intervals for each date.  All sites can be aggregated by sample dates or by "location groups" in the site metadata file.  For example, all sites along the same river or tributary can be aggregated.  The following demonstrates an aggregation of sites by the mainstem of the Assabet river and its tributaries using the `group = "locgroup"` argument.  A user can specify any desired grouping in the `Location Group` column of the site metadata file.

```{r}
#| fig-height: 4
#| fig-width: 8
#| echo: true
#| out-width: '95%'
#| fig-align: 'center'
anlzMWRdate(fset = fsetls, param = 'DO', group = 'locgroup', 
  thresh = 'fresh')
```

The `anlzMWRsite()` function can be used to view data organized by site. This function summarizes results for a single parameter using boxplots or barplots separately for each site on the x-axis. Boxplots or barplots can also include overlaid jittered points of the observations or only the jittered points can be shown, as for the `anlzMWRseason()` function. Results at each site can also be grouped by the `Result Attribute` column in the results file, where a user can enter arbitrary grouping criteria for samples. For example, sites can be grouped by wet or dry conditions if this information is included in the results file to evaluate the effects of precipitation on water chemistry or pollutants.  Individual sites to plot can also be specified with the `site` argument, where the default is to plot all sites.

```{r}
#| fig-height: 4
#| fig-width: 8
#| echo: true
#| out-width: '95%'
#| fig-align: 'center'
anlzMWRsite(fset = fsetls, param = "DO", thresh = "fresh", type = "box", 
  site = c("ABT-026", "ABT-062", "ABT-077", "ABT-144", "ABT-237", "ABT-301"),
  resultatt = c('DRY', 'WET'), byresultatt = TRUE)
```

Finally, maps of results at each site can be created with the `anlzMWRmap()` function.  Parameter values at each site are summarized with colors indicating relative values and the maps can include various spatial information for context. For example, hydrologic lines and waterbodies from the National Hydrograpy Dataset (NHD) can be shown with varying level of detail and base maps can be included from the *ggmap* package [@Kahle13].  NHD data were included to provide more specific information on hydrologic features of interest because of insufficient detail provided by standard basemaps.  These datasets are available from an external source and clipped to an approximate bounding box for the selected stations.  Currently, only flowlines and waterbodies that intersect Massachusetts are included, although examples are provided in a package [vignette](https://massbays-tech.github.io/MassWateR/articles/modifying.html#modifying-masswater-plots) that demonstrate how to include custom shapefiles in the map.  The below example shows mean dissolved oxygen concentrations using the medium level of detail for the NHD maps (`addwater = "medium"`).

```{r}
#| fig-height: 5
#| fig-width: 5
#| echo: true
#| out-width: '95%'
#| fig-align: 'center'
anlzMWRmap(fset = fsetls, param = "DO", addwater = "medium")
```

@@

# Building a community and future work

@@

# Summary

@@

# Acknowledgments

This work was supported by an Exchange Network grant from the US Environmental Protection Agency awarded to the Massachusetts Bays National Estuary Partnership, Grant No\. OS\-84029801\-0. We thank the early users of *MassWateR* that contributed ideas on improving the package during initial testing.  We are grateful for discussions with the Massachusetts Department of Environmental Protection that guided the presentation of results in the quality control assessments and WQX staff at USEPA for consult on submitting data formatted with *MassWateR*.

# References

>>>>

knitr:
  opts_chunk: 
    fig.path: 'images/'

execute: 
  echo: false
  warning: false
  message: false
  ft.keepnext: false
  ft.align: left
  fig-dpi: 300
---

```{r}
#| include: false
#|

library(flextable)
library(dplyr)
library(here)

flextable::set_flextable_defaults(font.size = 7, padding = 0)

@@

Short title: *MassWateR* R package for water quality data

@@

The long-term protection and restoration of aquatic resources depends on robust monitoring data; data that require systematic quality control and analysis tools. The *MassWateR* package facilitates quality control, analysis, and data sharing for discrete surface water quality data collected by monitoring programs of various size and technical capacity. The tools were developed to address regional needs for programs in Massachusetts, USA, but the principles and outputs can be applicable to monitoring data collected anywhere.  Users can create quality control reports, perform outlier analyses, and assess trends by season, date, and site for more than 40 parameters. Users can also prepare data for submission to the United States Environmental Protection Agency Water Quality Exchange, thus sharing data to the largest water quality database in the United States. The automated and reproducible workflow offered by *MassWateR* is expected to increase the quantity and quality of publicly available data to support the management of aquatic resources.

@@

![Workflow demonstrating how a user could engage with the *MassWateR* package. WQX: Water Quality Exchange; QC: Quality Control.](images/workflow.png){#fig-workflow width=100% fig-alt="Workflow diagram showing four ways to engage with *MassWateR*"}

@@

![Pseudocode demonstrating the iterative process of importing a required data file for *MassWateR*.  All read functions import an Excel file from the user-specified path (e.g., `respth`) and the imported file is then passed to a check function.  The function exits if an error is encountered, allowing the user to manually fix the identified error and then import again.  After all checks are passed, a formatting function is applied to correct minor issues (e.g., standardize date format as YYYY-MM-DD) and the final data object is returned (e.g., `resdat`).](images/importflow.png){#fig-importflow fig-alt="Diagram showing pseudocode for importing and fixing a data file with *MassWateR*" width=100%}

@@

The `anlzMWRoutlier()` function uses the results file and data quality objectives for accuracy to plot potential outliers for a parameter by month.  The accuracy file is used to automatically identify the y-axis scaling (as arithmetic or log) and to replace concentrations with appropriate values for those labelled as beyond detection. Outliers are defined using the standard definition of 1.5 times the interquartile range (the 25th to 75th percentile) of a parameter and can be visually identified as points above or below the whiskers in the boxplots.  The station name for a point is also shown for identification.  The `param` argument specifies the water quality parameter to evaluate and the `group` argument specifies how the boxplots are grouped (by year, month, week, or station).  In the following example (@fig-outlier), three stations are identified as potential outliers for the respective month.  Note that the outliers may differ slightly by the grouping choice as they are specific to each value on the x-axis.

```{r}
#| label: fig-outlier
#| fig-cap: "Example plot showing identification of outliers in a water quality dataset.  The plot was created using the `anlzMWRoutlier()` function."
#| fig-height: 4
#| fig-width: 8
#| echo: true
#| out-width: '95%'
#| fig-align: 'center'
anlzMWRoutlier(fset = fsetls, param = "DO", group = "month")
```

@@

![](images/qcreview1.png)

![](images/qcreview2.png)

@@

Seasonal trends can be evaluated using the `anlzMWRseason()` function that summarizes results for a single parameter using boxplots or barplots with seasonal groups assigned to months or weeks of the year (@fig-season). Boxplots or barplots can also include jittered points of the observations on top or only the jittered points can be shown.  @fig-season demonstrates a jittered boxplot of dissolved oxygen observations by month.

```{r}
#| label: fig-season
#| fig-height: 4
#| fig-width: 8
#| echo: true
#| out-width: '95%'
#| fig-align: 'center'
#| fig-cap: "Example plot showing water quality results grouped by month.  The plot was created using the `anlzMWRseason()` function."
anlzMWRseason(fset = fsetls, param = "DO", thresh = "fresh", 
  group = "month", type = "jitterbox")
```

@@

The `anlzMWRdate()` function plots results as time series on the date the samples or measurements were collected (@fig-date).  This information can be used to evaluate how samples have changed at individual sites over time or as aggregate samples across sites.  In the former case, individual points and lines are used for each site with appropriate labels. For the latter case, sites are aggregated and summary statistics are shown as the mean with optional 95% confidence intervals for each date.  All sites can be aggregated by sample dates or by "location groups" in the site metadata file.  For example, all sites along the same river or tributary can be aggregated.  @fig-date shows an aggregation of sites by the mainstem of the Assabet river and its tributaries using the `group = "locgroup"` argument.  A user can specify any desired grouping in the `Location Group` column of the site metadata file.

```{r}
#| label: fig-date
#| fig-height: 4
#| fig-width: 8
#| echo: true
#| out-width: '95%'
#| fig-align: 'center'
#| fig-cap: "Example plot showing water quality results grouped by sample date.  The plot was created using the `anlzMWRdate()` function."
anlzMWRdate(fset = fsetls, param = 'DO', group = 'locgroup', 
  thresh = 'fresh')
```

The `anlzMWRsite()` function can be used to view data organized by site (@fig-site). This function summarizes results for a single parameter using boxplots or barplots separately for each site on the x-axis. Boxplots or barplots can also include overlaid jittered points of the observations or only the jittered points can be shown, as for the `anlzMWRseason()` function. Results at each site can also be grouped by the `Result Attribute` column in the results file, where a user can enter arbitrary grouping criteria for samples. For example, sites can be grouped by wet or dry conditions if this information is included in the results file to evaluate the effects of precipitation on water chemistry or pollutants.  Individual sites to plot can also be specified with the `site` argument, where the default is to plot all sites.

```{r}
#| label: fig-site
#| fig-height: 4
#| fig-width: 8
#| echo: true
#| out-width: '95%'
#| fig-align: 'center'
#| fig-cap: "Example plot showing water quality results grouped by site.  The plot was created using the `anlzMWRsite()` function."
anlzMWRsite(fset = fsetls, param = "DO", thresh = "fresh", type = "box", 
  site = c("ABT-026", "ABT-062", "ABT-077", "ABT-144", "ABT-237", "ABT-301"),
  resultatt = c('DRY', 'WET'), byresultatt = TRUE)
```

Finally, maps of results at each site can be created with the `anlzMWRmap()` function (@fig-map).  Parameter values at each site are summarized with colors indicating relative values and the maps can include various spatial information for context. For example, hydrologic lines and waterbodies from the National Hydrograpy Dataset (NHD) can be shown with varying level of detail and base maps can be included from the *ggmap* package [@Kahle13].  NHD data were included to provide more specific information on hydrologic features of interest because of insufficient detail provided by standard basemaps.  These datasets are available from an external source and clipped to an approximate bounding box for the selected stations.  Currently, only flowlines and waterbodies that intersect Massachusetts are included, although examples are provided in a package [vignette](https://massbays-tech.github.io/MassWateR/articles/modifying.html#modifying-masswater-plots) that demonstrate how to include custom shapefiles in the map.  @fig-map shows mean dissolved oxygen concentrations using the medium level of detail for the NHD maps (`addwater = "medium"`).

```{r}
#| label: fig-map
#| fig-height: 5
#| fig-width: 5
#| echo: true
#| out-width: '95%'
#| fig-align: 'center'
#| fig-cap: "Example map showing water quality results averaged by sampling location. The map was created using the `anlzMWRmap()` function."
anlzMWRmap(fset = fsetls, param = "DO", addwater = "medium")
```

@@

# Discussion

@@

# Conclusions

@@

# Acknowledgments {.unnumbered}

We thank the early users of *MassWateR* that contributed ideas on improving the package during initial testing.  We are grateful for discussions with the Massachusetts Department of Environmental Protection that guided the presentation of results in the quality control assessments and WQX staff at USEPA for consult on submitting data formatted with *MassWateR*.

# References {.unnumbered}
